Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs

### 2019-07-02 manuscript/manuscript.Rmd
Added citations

---
title: "Long-term Water Quality Change in Rhode Island Lakes and Ponds"
output: 
  word_document:
    reference_docx: manuscript_template.docx   
author:
- "Hollister. J. W. ^\\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A ^*2*^"
- "^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic Ecology Division, Narragansett, RI 02882"
- "^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston, RI 02881"
- "^*3*^ ORISE, Narragansett, RI 02882"
- "^\\*^  *corresponding author: hollister.jeff@epa.gov*"
bibliography: manuscript.bib
csl: plos.csl
abstract: The University of Rhode Island’s Watershed Watch Volunteer Monitoring Program has been collecting water quality data on Rhode Island lakes and ponds for close to three decades, allowing us to explore long-term trends in common water quality parameters.  Not all lakes and ponds in the study area were sampled across the full time period and lakes were often added in geographic clusters (e.g. in urbanized northern Rhode Island).  Not unlike how long-term temperature records are analyzed, we centered and scaled (i.e., the z-score) water quality measurements on a per-station basis. This provides a robust and commonly scaled measurement to explore this data for long-term trends.  State-wide aggregation shows increasing temperature, chlorophyll *a*, and total nitrogen. Interestingly, total phosphorus is showing a decline, perhaps reflecting the management focus on phosphorus reductions.  While yearly trends are useful, they do mask month-to-month variability differences across sites.  Additionally, while most sites track the yearly trend in decreasing water quality, there are bright spots with some sites improving over the 25 years. Contrary to previously reported analyses that show relatively stable water quality at the regional scale, our analysis shows that long-term water quality trends within Rhode Island show some parameters improving while others are in decline.  Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for identifying trends in environmental condition. 
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(captioner)
source(here::here("R/functions.R"), encoding = "UTF-8")
opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
tab_num <- captioner(prefix = "Table")
fig_num <- captioner(prefix = "Figure")
```

@@

1. Water quality in last several decades (Why Water Quality Paragraph(s))
  - Why of interest
    - cyanos stuff
    - CWA stuff
  - Understand long-term trends from local, to regional, to national
  - What variables are of interest and why
    - productivity
    - climate change
    - Nutrient enrichment stuff

2. Temporal and Spatial scaling of water quality trends Paragraph

3. Work has been done in the larger Northeastern Region of US (Northeastern Paragraph)
  - Shown to be stable in LAGOS (Maybe other studies in NE)
  - "Given what is known about long term trends in water quality within the broader region of the northeastern United States [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds."

4. Rhode Island Paragraph (May be combined with #3)

4. Thus, the goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island and answer two particular questions.  First, what are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and lake temperature and second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.   In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends>.

@@

The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and pond sampling sites")`).  The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and ponds for over 25 years. The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URI WW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy. Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and ponds, weekly secchi depth and water temperature are recorded, along with bi-weekly Chl-a and dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory for nutrients, alkalinity, pH, and bacteria.

*`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and pond sampling sites")`*
![](../figures/ww_map.jpg)

In particular we selected data from the URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less. For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and ponds which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), and `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, temp")` had data for all 4 parameters.

Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].  The authors found little change in water quality trends across this region [@Oliver_2017] .  We repeated our analysis (see **Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].  Temperature data were not availble, thus we only examine trends, using our analytical methods, for Total Nitrogren, Total Phosphorus, and Chlorophyll *a* from the LAGOSNE database.

## Trend Analysis

There are many different methods for analyzing time series data for trends.  One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been addedd to a dataset.  For instance, if you have long-term data on water quality for several ponds, yet the cleaner ponds wer sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.  Thus, it is necessary to account for this type of site-specific variation.  This is similar to how the long-term temperature trends are analyzed using temperature anomolies [FILL IN GROUP DOING THIS].  The general approach is to calculate site-specific deviations from a long-term mean over pre-determined reference period.  This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change relative to the reference period.

@@

The anomaly approach is very useful and is the approach we chose to follow.  However, using anomalies works best with a single measure (e.g temperature) or with mulitple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season will range from approximately 15 degrees celsius to a high of 30 degrees celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values we used the common approach of dividing each anomlaly by the standard deviation for the reference period.  This value is commonly referred to as a Z-score.  We use these z-scores to examine each water quality variable for a trend over the time period of 1990 to 2016.

Furthermore, since we are interest in water quality trends over time, we wanted to explore how each site was responding.  Thus, our z-scores were calculated over the reference period, 1990-2016, for each site.  In other words the long-term mean and standard deviation were calculated on a site-by-site basis.

### Kendall's Tau

Kendall's Tau and the Mann-Kendall test are commonly used to test for monotonic trends water quality data [@mann1945nonparametric; @helsel2002statistical; @hipel1994time].  We followed these standard procedures to test for postive or negative trends in lake temperature, chlorophyll *a*, total nitrogen, and total phosphorus.

@@

Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].  The data we used in our analysis, the University of Rhode Island's Watershed Watch data, was inlcuded in that, but we were curious if regional trends were masking local scale trends in Rhode Island.  In addition to the analysis condcuted by Oliver et al. [-@Oliver_2017], we analyzed the same dataset but using the approach outlined here.

@@

During the sampling period of 1990 to 2016 Rhode Island lakes and ponds average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).

*`r tab_num("ww_params")`*

```{r ww_params, echo=FALSE, results="asis"}
knitr::kable(ww_param_ranges, format = "markdown")
```

For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).

*`r tab_num("lagos_params")`*

```{r lagos_params, echo=FALSE, results="asis"}
knitr::kable(lagos_param_ranges, format = "markdown")
```

@@

Test test test

Average yearly scaled temperature in lakes and ponds appear to be increasing (Kendall's Tau: `r round(ww_temp_kt$tau,3)` , p-value: `r round(ww_temp_kt$sl, 4)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (Kendall's Tau: `r round(ww_chla_kt$tau,3)` , p-value: `r round(ww_chla_kt$sl, 4)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).

*`r fig_num(name = "temp_ri")`*

![](../figures/ww_temp_trends.jpg)

*`r fig_num(name = "chla_ri")`*

![](../figures/ww_chla_trends.jpg)

```{r no_early_tn, echo=FALSE}
ww_tn_kt_no9394 <- Kendall(ww_tn_df[c(-1,-2),]$year, ww_tn_df[c(-1,-2),]$mn_value)
```

Average yearly trends for nutrients showed weaker or no trends over time.  Total nitrogen did have a significant trend (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).  Since 1995, the yearly trend is not significant (Kendall's Tau: `r round(ww_tn_kt_no9394$tau,3)` , p-value: `r round(ww_tn_kt_no9394$sl, 4)`).  Total phosphorus does not show a trend over time in the yearly scaled values (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).

*`r fig_num(name = "tn_ri")`*

![](../figures/ww_tn_trends.jpg)

*`r fig_num(name = "tp_ri")`*

![](../figures/ww_tp_trends.jpg)

@@

(`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).

*`r fig_num(name = "tn_lagos")`*

![](../figures/lagos_tn_trends.jpg)

(`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`).

*`r fig_num(name = "tp_lagos")`*

![](../figures/lagos_tp_trends.jpg)

(`r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).

*`r fig_num(name = "chla_lagos")`*

![](../figures/lagos_chla_trends.jpg)

>>>>

---
title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management. Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine its effectiveness requires long-term datasets.  A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States, however, trends at smaller scales may be masked.  We address this by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.  URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. These data, included in LAGOSNE, allow for comparison of water quality trends at regional and state extents. We assess for trends with z-scores (i.e. scaled anomalies) calculated on a per-station basis and examine yearly averages.  Temperature and chlorophyll *a* are increasing.  Total nitrogen (TN) shows a weak increasing trend driven by low years in the early 1990s.  Total phosphorus (TP) and the total nitrogen:total phosphorus ratio (TN:TP) are stable.  Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP all stable over time.  In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes. Although not causal, this analysis suggests an association between lake temperature and productivity.  Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends. '
---

```{r setup, include=FALSE}
source(here::here("R/functions.R"), encoding = "UTF-8")
opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
tab_num <- captioner(prefix = "Table")
fig_num <- captioner(prefix = "Figure")
options(scipen = 2, digits = 5)
```

@@

Aquatic ecosystems have been altered as the result of human activities modifying nutrient cycling on a global scale [@vitousek1997human; @finlay2013human; @filippelli2008global]. Because of their position in the landscape, lakes can function as integrators and sentinels for these anthropogenic effects [@williamson2008lakes; @schindler2009lakes]. Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].  This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate]. To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication]. These contrasting studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales.

Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis. These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing]. Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental]. Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking]. These trophic state models indicate that landscape variables (i.e. ecoregion, elevation, and latitude) are important and that regional trends exist. Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance]. Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].

Previous studies using regional data from the northeastern and midwestern United States have investigated spatial and temporal water quality trends and have shown differences based on scale. Macro-scale (i.e. subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale]. This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake]. On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence]. Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017]. Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality in Rhode Island lakes and reservoirs.

Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).   URI’s Watershed Watch is a citizen science program founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.  Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.  These efforts have been ongoing in some waterbodies since 1988.  These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing].  Additionally, the fact that volunteer monitors are used has allowed for this program to persist much longer than agency run programs that are often prone to uncertain funding.  Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al. [@Oliver_2017], but also allowed us to examine the long-term trends in Rhode Island.

Thus, the goals of this study were to look at approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.  First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature. Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states. Another focus of this paper was to apply existing methodologies for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends. Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].

@@

The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).  The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and reservoirs for over 25 years. The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy. Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory.

![`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`](../figures/ww_map.jpg)

In particular we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less. For this analysis we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  For each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.

Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].  The authors found little change in water quality trends across this region [@Oliver_2017] .  We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].  Temperature data were not available, thus we only examine trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.

## Water Quality Trend Analysis

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites. For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies then a simple comparison of raw-values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies (e.g., [@jones1996calculating]).  The general approach is to calculate site-specific deviations from a long-term mean over a pre-determined reference period.  This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change over the specified reference period.

@@

Anomaly analysis works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).  The resultant value is commonly referred to as a z-score.  We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.   Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.  We refer to this approach as the site-specific z-scores.

### Summarizing site-specific z-scores

Methods for calculating the site-specific z-scores and the yearly averages are as follows and are presented graphically in `r fig_num(name = "z_schematic", caption = "Example calculation of the site-specific z-scores and yearly mean z-scores.", display = "cite")`.  Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("z_schematic", display = "cite")` is available from <fill in with zenodo link> [@zenodo_citation].

![`r fig_num("z_schematic")`](../figures/z-score_schematic.jpg)

The general steps, outlined in `r fig_num("z_schematic", display = "cite")` and listed below, are repeated for each of the water quality parameters.

1. For each site, calculate the annual means, producing a single mean value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site means across years are assumed to be independent.
2. Calculate the long-term reference mean and reference standard deviation for all sites. This results in a single long-term mean and standard deviation for each of the sites.
3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

### Linear regression on annual means

Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g. $\beta$~1~ = 0).

@@

Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].  While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.  The analysis conducted by Oliver et al. [-@Oliver_2017] is a robust approach; however, to make direct comparisons between Rhode Island and the region,  we re-analyzed the same dataset used by Oliver et al. [-@Oliver_2017] but using the trend analysis approach outlined above.

@@

During the period of 1990 to 2016 Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).

```{r ww_params, echo=FALSE, results="asis"}
knitr::kable(ww_param_ranges, format = "markdown")
```
*`r tab_num("ww_params")`*

For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).

```{r lagos_params, echo=FALSE, results="asis"}
knitr::kable(lagos_param_ranges, format = "markdown")
```
*`r tab_num("lagos_params")`*

@@

Average annual scaled temperature in lakes and reservoirs appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).  Panel A. Yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.  Panel B. Yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).

![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends.jpg)

![`r fig_num(name = "chla_trend")`](../figures/paired_chla_trends.jpg)

```{r no_early_tn, echo=FALSE}
ww_tn_rl_no9394 <- lm(mn_value~year, data = ww_tn_df[c(-1,-2),]) %>%
  tidy() %>%
  slice(2) %>%
  select(slope = estimate, p.value)
```

Mean annual trends for nutrients were weaker or showed no trend over time. The data suggest a positive trend in TN (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average z-score). Panel A. Yearly averaged TN z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).  Since 1995, the yearly trend is shows much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).  TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).  Panel A. Yearly averaged TP z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TP z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.). The pattern is the same for the TN:TP ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average z-score).  Panel A. Yearly averaged TN:TP ratio z-scores from the URI Watershed Watch dataset.  Panel B. Yearly averaged TN:TP ratio z-scores from the LAGOSNE dataset. Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).  Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from <link to zenodo> [@zenodo_citation].

![`r fig_num(name = "tn_trend")`](../figures/paired_tn_trends.jpg)

![`r fig_num(name = "tp_trend")`](../figures/paired_tp_trends.jpg)

![`r fig_num(name = "np_trend")`](../figures/paired_np_trends.jpg)

@@

In general, there was little to suggest broad regional changes. Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP ratio was also flat (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
